---
title: "Lab 5"
author: "Sawyer Balint"
date: "Fall 2024; Marine Semester Block 3"
output:
  bookdown::pdf_document2: default
---

```{r setup, include=FALSE}
rm(list = ls()) #clear environment
knitr::opts_chunk$set(echo = TRUE)
library(bookdown)
```

# Introduction

This document is available at https://github.com/sjbalint/BI521/tree/main/scripts/labs

```{r, message=FALSE}
#import packages
library(tidyverse) #for data wrangling
library(here) #for filepath management
library(ggsci) #for colors
library(scales) #for log axis breaks
library(bbmle) #for stats

#custom graphing theme

update_geom_defaults("point", list(shape = 21, fill="grey", stroke=0.8))

theme <- list(
  theme_classic(),
  scale_color_viridis_d(option="inferno", end=0.8),
  scale_fill_viridis_d(option="inferno", end=0.8),
  theme(legend.position="right"),
  labs(x="Time")
)

```

# Part I

## Task 1

```{r}

#function for S-Ricker model
gR_SRicker <- function(S,mu){
  with(mu,S^(1+delta)*exp(r*(1-S/K)))
}

SRicker_sim <- function(S0, mu, n_iter){
  
  R0 <- gR_SRicker(S0, mu)
  
  df <- data.frame(S=S0, R=R0, t=0)
  
  #empty list to store results
  result.list <- list(df)
  
  i_model <- 1
  
  while(i_model<n_iter){
    
    i_model <- i_model + 1
    
    R <- gR_SRicker(S0, mu)
    S <- with(mu, R + Sa*S0)
    
    df <- data.frame(S=S, R=R, t=i_model)
    
    S0 <- S
    R0 <- R
    
    #log results
    result.list <- c(result.list, list(df))
  }
  
  #compile results
  result.df <- bind_rows(result.list)
  
  return(result.df)
  
}

#try it out
mu <- list(r=2.6, K=3/4, delta=1, Sa=0.6)
df <- SRicker_sim(S0=0.04, mu=mu, n_iter=50)

#looks good
head(df)

#loop over a range of S0
S0.list <- seq(0.01,0.1, by=0.01)

result.list <- list()

for (S0_i in S0.list){
  
  df <- SRicker_sim(S0=S0_i, mu=mu, n_iter=20) %>%
    mutate(S0 = S0_i)
  
  result.list <- c(result.list, list(df))
  
}

result.df <- bind_rows(result.list) %>%
  mutate(S0 = as.factor(S0))

ggplot(result.df, aes(t,S,color=S0, group=S0))+
  theme+
  geom_line()+
  geom_hline(yintercept=0, linetype="dashed")

```

> **Fig. 1:** Sigmodal Ricker with *S_0* values ranging from 0.01 to 0.1 (color). We find that S_0 equal to 0.01, 0.02, and 0.03 result to a population that crashes to zero.

## Task 2

```{r}

#new function that includes fishing mortality
SRicker_Fm_sim <- function(S0, mu, Fm, n_iter){
  
  R0 <- gR_SRicker(S0, mu)
  
  df <- data.frame(S=S0, R=R0, t=0)
  
  #empty list to store results
  result.list <- list(df)
  
  i_model <- 1
  
  while(i_model<n_iter){
    
    i_model <- i_model + 1
    
    R <- gR_SRicker(S0, mu)
    S <- with(mu, (R + Sa*S0)*(1-Fm))
    
    df <- data.frame(S=S, R=R, t=i_model)
    
    S0 <- S
    R0 <- R
    
    #log results
    result.list <- c(result.list, list(df))
  }
  
  #compile results
  result.df <- bind_rows(result.list)
  
  return(result.df)
  
}

Fm.list <- seq(0,0.8, by=0.1)
S0=0.5

result.list <- list()

for (Fm_i in Fm.list){
  
  df <- SRicker_Fm_sim(S0=S0, mu=mu, Fm=Fm_i, n_iter=20) %>%
    mutate(Fm = Fm_i)
  
  result.list <- c(result.list, list(df))
  
}

result.df <- bind_rows(result.list) %>%
  mutate(Fm = as.factor(Fm))

#make a plot
ggplot(result.df, aes(t,S,color=Fm, group=Fm))+
  theme+
  geom_line()+
  geom_hline(yintercept=0, linetype="dashed")
```

> **Fig. 2:** Higher fishing mortality results in a lower equilibrium population level and a longer time to reach equilibrium. At mortalities of 0.6, 0.7, and 0.8, the population crashes to zero.

> This simulation implements a constant effort model of fishing, where the yield is dependent on the size of the population.

## Task 3

```{r}
#function for S-Ricker model with noise
gR_SRicker_noise <- function(S,mu){
  with(mu,rlnorm(1, meanlog=(1+delta)*log(S)+(r*(1-(S^(1))/K)), sdlog=tau_r))
}

#new function that includes noise
SRicker_Fm_noise_sim <- function(S0, mu, Fm, n_iter){
  
  R0 <- gR_SRicker_noise(S0, mu)
  
  df <- data.frame(S=S0, R=R0, t=0)
  
  #empty list to store results
  result.list <- list(df)
  
  i_model <- 1
  
  while(i_model<n_iter){
    
    i_model <- i_model + 1
    
    R <- gR_SRicker_noise(S0, mu)
    S <- with(mu, (R + Sa*S0)*(1-Fm))
    
    df <- data.frame(S=S, R=R, t=i_model)
    
    S0 <- S
    R0 <- R
    
    #log results
    result.list <- c(result.list, list(df))
  }
  
  #compile results
  result.df <- bind_rows(result.list)
  
  return(result.df)
  
}

S0=0.2

result.list <- list()

for (Fm_i in seq(0,0.6, by=0.2)){
  
  for (taur_r_i in c(0,0.1)){
  mu <- list(r=2.6, K=3/4, delta=1, Sa=0.6, tau_r=taur_r_i)
  
  df <- SRicker_Fm_noise_sim(S0, mu, Fm=Fm_i, n_iter=20) %>%
    mutate(tau_r=taur_r_i,
           Fm=Fm_i)
  
  result.list <- c(result.list, list(df))
  
  }
  
}

result.df <- bind_rows(result.list) %>%
  mutate(tau_r=factor(tau_r),
         Fm = factor(Fm))

#make a plot
ggplot(result.df, aes(t,S,color=Fm, linetype=tau_r))+
  theme+
  geom_line()+
  geom_hline(yintercept=0, linetype="dashed")

```

> **Fig. 3:** I modified S0 and the range of fishing mortality tested. The additional "noise" of *tau_r* = 1 is observed in the dashed lines.

> This version is including process error - variations in the actual population of the fishery, rather than noise stemming from incomplete counting.

# Model Comparison

## Task 4

```{r}
#make functions
gR_allee <- function(S,mu){
  with(mu,rlnorm(1, meanlog=(1+delta)*log(S) + (r*(1-(S^(1))/K)), sdlog=tau_r))
}
gS_allee <- function(S,Fm,mu){
  with(mu, (gR_allee(S,mu)+Sa*S)*(1-Fm))
}

gR_ricker <- function(S,mu){
  with(mu, rlnorm(1, meanlog=(1)*log(S) + (r*(1-(S^(1))/K)), sdlog=tau_r))
}
gS_ricker <- function(S,Fm,mu){
  with(mu,(gR_ricker(S,mu) + Sa*S)*(1-Fm))
}

simulation <- function(S, Fm, dFm, mu, n_burn, n_iter){
  
  R <- gR_allee(S,mu)
  
  result.list <- list()
  
  for (model_i in 1:(n_iter+n_burn)){
    
    R <- gR_allee(S,mu)
    S <- with(mu, (R+Sa*S)*(1-Fm))
    
    Fm <- Fm + dFm
    
    df <- data.frame(R=R, S=S, Fm=Fm, t=model_i)
    
    result.list <- c(result.list, list(df))
    
  }
  
  result.df <- bind_rows(result.list)
  
  return(result.df)

}

#run the simulation
mu <- list(r=2.1, K=1, delta=0.6, Sa=0.65, tau_r=0.15)
df <- simulation(S=0.8, Fm=0, dFm=0.005, mu=mu, n_burn=50, n_iter=150)

plot.df <- df %>%
  pivot_longer(!t)

#visualize the simulation results
ggplot(plot.df, aes(t, value))+
  theme+
  geom_line()+
  geom_hline(yintercept=0, linetype="dashed")+
  facet_wrap(.~name, ncol=1, strip.position="left", scales="free_y")+
  theme(strip.placement="outside",
        strip.background=element_blank())+
  labs(y=NULL)

#make two models with different deltas
data_fit_1 <- data.frame(time=61:110, S=df$S[61:110], R=df$R[62:111])

mle2.m() <- mle2(R ~ dlnorm((1+0)*log(S)+alpha-beta*(S^(1+0)), sdlog=tau_r),
                 method="BFGS",
                 control=list(maxit=5000),
                 data=data_fit_1,
                 start=list(alpha=2.5, beta=1/2.2, tau_r=0.1))

```

> **Fig. 4:**

# Extension



# Reflection