---
title: "Splines and over-fitting"
output:
  pdf_document: default
  html_document: default
date: "2023-11-14"
---

```{r setup, include=FALSE}
library(tidyverse)
```

## R Markdown

Cell growth rate (per day) in diatom of genus Thalassiosira taken from Montagnes and Franklin 2001.

```{r cars}


f_temp <- function(sst){rlnorm(1,meanlog=log(0.014*sst + 0.25),sdlog = 0.1)}
set.seed(2003)

x_temp <- runif(30,min=10,max=20)
y_growth_rate <- vapply(x_temp,f_temp,FUN.VALUE = 0)

df_growth <- data.frame(temp=x_temp,growth_rate=y_growth_rate)
```

## Including Plots

Let's just consider the first 10 data points and assume that's our initial data set.

```{r pressure, echo=FALSE}
df_growth %>% 
  filter(row_number() <= 10) %>% 
  ggplot(aes(x=temp,y=growth_rate)) + 
  geom_point(shape=1) +
  theme_bw() +
  labs(x="Temperature (ºC)",y=bquote(Growth~Rate~(mu~day^-1)))
```

Fit a linear regression.

```{r}

```



Fit a spline.

```{r}
?smooth.spline
```


There's a lot of ways R let's you control how this procedure works; to be honest I'm not clear on how all these operate. The original notions of splines worked on "knots", though.

```{r}
spline_5ks_0spar <- smooth.spline(x_temp[1:10],y_growth_rate[1:10],all.knots = seq(0,1,by=.25),spar=0.0)
spline_5ks_1spar <- smooth.spline(x_temp[1:10],y_growth_rate[1:10],all.knots = seq(0,1,by=.25),spar=1.0)

plot(predict(spline_5ks_0spar,seq(10,20,by=0.1)),type="l",col="blue",,ylim=c(0.2,.7)) +
  lines(predict(spline_5ks_1spar,seq(10,20,by=0.1)),col="lightgreen") +
  points(x_temp[1:10],y_growth_rate[1:10])
```

```{r}
spline_10ks_0spar <- smooth.spline(x_temp[1:10],y_growth_rate[1:10],all.knots = seq(0,1,by=.1),spar=0)
spline_10ks_1spar <- smooth.spline(x_temp[1:10],y_growth_rate[1:10],all.knots = seq(0,1,by=.1),spar=1)

plot(predict(spline_10ks_0spar,seq(10,20,by=0.1)),type="l",col="blue",ylim=c(0.2,.7)) +
  lines(predict(spline_10ks_1spar,seq(10,20,by=0.1)),col="lightgreen") +
  points(x_temp[1:10],y_growth_rate[1:10])
```

```{r}
spline_auto <- smooth.spline(x_temp[1:10],y_growth_rate[1:10])

print(spline_auto)
```
```{r}
plot(predict(spline_auto,seq(10,20,by=0.1)),type="l",col="lightgreen")
abline(a=0.25,b=0.014)
```
