---
title: "DEMO - While and if"
author: "ER Deyle"
date: "Fall 2023; Marine Semester Block 3"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



We've been talking about derivatives, equilibria, and finding zeros, so the examples in this demo will use those.

In other programming languages, *if* statements are sometimes called *if-then* statements, since the syntax explicitly includes "then". There's also the opportunity to include an alternative piece of code for the cases where the conditional statement is "FALSE". This would constistute and *if-else* statement.

## Conditionals

There are a lot of relational operators in R, including `>`, `<`, `==`, `!=`, `>=`, and `<=`. Try these out.

`>`: "Is LHS greater than RHS?"

```{r}
1 > 2
```

`<`: "Is LHS greater than RHS?"

```{r}
1 < 2
```

`==`: "Is LHS equal to RHS?"

```{r}
1 == 2
```

`!=`: "Is LHS not equal to RHS?"

```{r}
1 != 2
```

`>=`: "Is LHS greater than or equal to RHS?"

```{r}
1 >= 2
```

`<=`: "Is LHS less than or equal to RHS?"

```{r}
1 <= 2
```

There are a lot of other functions in R that return a logical (i.e. a `TRUE` or `FALSE` value). Here are some useful ones to check a single variable, vector, or list: `is.empty()`, `is.null()`, `is.na()`, `is.finite()`. There are also a lot of useful set-operations for conditionals, like `%in%`.

```{r}
3 %in% 1:10
```

```{r}
13 %in% 1:10
```

```{r}
"d" %in% list("a","b","c","d","e")
```

## `If`

The `If` construction in R bears some structural similarity to `for` loops. The general syntax is:

```{r,eval=FALSE}
if(CONDITION){
  CODE
}
```

The enclosed "CODE" will only evaluate if `CONDITION` is `TRUE`. So, if we simply put the value `TRUE` in the arguement of `if()`, the code will evaluate:

```{r}
if(TRUE){
  print("Yes!")
}
```

The code will also evaluate if we put an expression in that evaluates to `TRUE`.

```{r}
if(1 < 2){
  print("Yes!")
}
```

This statement can include variables, but of course they need to have defined values already.

```{r}
a <- 1
b <- 2

if(a < b){
  print("Yes!")
}
```

When the `CONDITION` is `FALSE`, then R will not evalute the code. For example, if we simply put the value `FALSE` in the arguement of `if()`:

```{r}
if(FALSE){
  print("Yes!")
}
```

Nothing happened. And as before, the code will also be skipped if we put in an expression that evaluates to `FALSE`.

```{r}
if(1 > 2){
  print("Yes!")
}
```


## `Else`

As I mentioned above, it's possible to present an alternative to the code enclosed in the `{}` using `else`. The general syntax is as follows:

```{r,eval=FALSE}
if(CONDITION){
  CODE_1
}else{
  CODE_2
}

```

When the `CONDITION` is `TRUE`, the code in the first set of braces (`CODE_1`) is evaluated. Alternatively, when the `CONDITION` is `FALSE`, the code in the second set of braces (`CODE_2`) is evaluated.

```{r}
a <- 1
b <- 2

if(a < b){
  print("Yes!")
}else{
  print("No!")
}
```

```{r}
a <- 3
b <- 2

if(a < b){
  print("Yes!")
}else{
  print("No!")
}
```

**EXERCISE** Write an `if` statement to check if a variable is an integer. Use the function `round()` (type `?round`).

```{r}
# YOUR CODE HERE
```

## `while`.

In some sense, `while` loops are a blend of `for` and `if`. The basic construction is as follows:

```{r,eval=FALSE}
while(CONDITION){
  CODE
}
```

R will loop through the `CODE` enclosed in by the `{}` so long as `CONDITION` is `TRUE`. This immediately brings up an important point: if nothing ever changes the `CONDITION`, then it's possible a `while` loop will never finish on its own. When we started with `if` above, we first just set the `CONDITION` to be `TRUE`. Now when we do that, we will have to force R to suspend calculation or it will just keep looping.

```{r,eval=FALSE}
while(TRUE){
  print("1 second has passed")
  Sys.sleep(1)
}
```

Let's do something constrained instead, like counting. We will start at 1, print the value, then add 1, and tell R to stop at 10.

```{r}
counter <- 1

while(counter < 10){
  print(counter)
  counter = counter + 1
}
```

Notice that as soon as counter got to 10, R stopped executing the code. If we wanted R to do the 10th iteration, we could do a few things like changing the conditional slightly:

```{r}
counter <-  1

while(counter <= 10){
  print(counter)
  counter = counter + 1
}
```
We could also be clever about when we check the value of `counter` versus when we print the value of `counter`. If we want to print the value first,

```{r}
counter <- 0

while(counter < 10){
  counter = counter + 1
  print(counter)
}
```

We can go back to one of our population models for a fish population and use a `while` loop instead!

```{r}
N_fish_t <- .5
dt <- 0.05

counter <- 0
while(counter < 40){
  counter <- counter + 1
  N_fish_t <- N_fish_t + (N_fish_t * (1 - N_fish_t))*dt
  print(N_fish_t)
}
```


Why would we want to? Well there's at least one convenient thing here- we can keep track of the actual time value in the model instead of just the number of steps, so our `while` condition will become

```{r}
N_fish_t <- .5
dt <- 0.05

t_model <- 0
while(t_model < 2){
  t_model <- t_model + dt
  N_fish_t <- N_fish_t + (N_fish_t * (1 - N_fish_t))*dt
  print(c(t_model,N_fish_t))
}
```

## Using `while`.


The real value of a `while` loop, however, is if you want to repeat a calculation a lot of times but you aren't sure exactly *how many times*. Recall before we discussed approximating the derivative using it's original definition:

$$ \frac{df}{dx} = lim_{\Delta x \rightarrow 0} \frac{f(x+\Delta x) - f(x)}{\Delta x} $$

How small do we need make the step size to get our numerical estimate of the derivative within a certain tolerance? There's no generic answer, but we can simulate it! Each step of the way, we reduce the step size by 50%.

```{r}
delta_x <- 1
tolerated_error <- 10^-5

f_cubic <- function(x) x^3 - 2*x + 1
dfdx_cubic <- function(x) 3*x^2 - 2

x_star <- 2

current_error <- ( f_cubic(x_star + delta_x) - f_cubic(x_star) )/delta_x - dfdx_cubic(x_star)

while(current_error > tolerated_error){
  
  # cut the step size by 2 before starting over
  delta_x <- delta_x/2
  current_error <- ( f_cubic(x_star + delta_x) - f_cubic(x_star) )/delta_x - dfdx_cubic(x_star)
  print(c(delta_x,current_error))

}
```
